Notes on Design and Implementation of Anglican

### 2015-07-28 paper/paper.tex
implementation highlights in progress
continuations

@@

\subsection{Literals}

@@

\subsection{Opaque Expressions}

@@

\subsection{General Expressions}

The most general form of CPS transformation receives an
expression and a continuation, and returns the expression
in CPS form with the continuation potentially called in multiple
tail positions. General expressions can be somewhat voluntarily
divided into several groups:
\begin{itemize}
	\item binding forms --- \texttt{let} and
		\texttt{loop/recur};
	\item flow control --- \texttt{if}, \texttt{when},
		\texttt{cond}, \texttt{case}, \texttt{and}, \texttt{or} and \texttt{do};
	\item function applications and \texttt{apply};
	\item probabilistic forms --- \texttt{predict},
		\texttt{observe}, \texttt{sample}, \texttt{store}, and \texttt{retrieve}.
\end{itemize}

>>>>

\input{anglican}

@@

\subsubsection{Literals}

@@

\subsubsection{Opaque Expressions}

@@

\subsubsection{General Expressions}

The most general form of CPS transformation receives an
expression and a continuation, and returns the expression
in CPS form with the continuation potentially called in multiple
tail positions. General expressions can be somewhat voluntarily
divided into several groups:
\begin{itemize}
	\item binding forms --- \texttt{let} and
		\texttt{loop/recur};
	\item flow control --- \texttt{if}, \texttt{when},
		\texttt{cond}, \texttt{case}, \texttt{and}, \texttt{or} and \texttt{do};
	\item function applications and \texttt{apply};
	\item probabilistic forms --- \texttt{predict},
		\texttt{observe}, \texttt{sample}, \texttt{store}, and \texttt{retrieve}.
\end{itemize}
Functions that transform general expressions accept the
expression and the continuation as parameters, and are
consistently named \texttt{cps-of-}\textit{form}, for example,
\texttt{cps-of-do}, \texttt{cps-of-store}.

\subsection{Implementation Highlights}

\subsubsection{Continuations}

Continuations are functions that are called in tail positions
with the computed value and state as their arguments --- in CPS
there is always a function call in every tail position and never
a value. Continuations are passed to CPS transformers, and when
transformers are called recursively, the continuations are
generated on the fly.

There are two critical issues related to generation of
continuations:
\begin{itemize}
	\item unbounded \textit{stack growth} in recursive code;
	\item code size \textit{explosion} when a non-atomic
		continuation is symbolically substituted in multiple
		locations.
\end{itemize}

In implementations of functional programming languages stack
growth is avoided through \textit{tail call optimization} (TCO).
However, Clojure does not support a general form of TCO, and
CPS-transformed code that creates deeply nested calls will
easily exhaust the stack. Anglican employs a workaround called
\textit{trampolining} --- instead of inserting a continuation
call directly, the transformer always wraps the call into a
\textit{thunk}, or parameterless function. The thunk is returned
and called by the trampoline (Clojure provides function
\texttt{trampoline} for this purpose) --- this way the
computation continues but the stack is collapsed on every
continuation call. Function \texttt{continue} implements the
wrapping.

To realize the potential danger of code size explosion, consider
CPS transformation of code
\begin{lstlisting}[frame=single,language=Anglican, basicstyle=\scriptsize\ttfamily, showstringspaces=false]
(if (adult? person)
  (if (male? person)
	 (choose-beer)
	 (choose_wine))
  (choose-juice))
\end{lstlisting}
with continuation 
\begin{lstlisting}[frame=single,language=Anglican, basicstyle=\scriptsize\ttfamily, showstringspaces=false]
(fn [choice _]
  (case (kind choice)
    :beer (beer-jar choice)
    :wine (wine-glass choice)
    :juice (juice-bottle choice)))
\end{lstlisting}
The code of continuation, represented by an \texttt{fn} form, will be
repeated three times. In general, CPS code can grow extremely
large if symbolic continuations are inserted repeatedly.

To circumvent this inefficiency, CPS transformers for
expressions with multiple continuation points (\texttt{if} and
derivatives, \texttt{and}, \texttt{or}, and \texttt{case}) bind
the continuation to a fresh symbol if it is not yet a symbol.
Macro \texttt{defn-with-named-cont} establishes the binding
automatically.

\subsubsection{Primitive Procedures}

\subsubsection{Probabilistic forms}

\subsubsection{Stochastic memoization}
